Reproducible Research: Peer Assessment 1
=============================================
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##Step 1: Loading and preprocessing the data
Using the "read.csv" function:
```{r echo=TRUE}
data<-read.csv("activity.csv")
```
A first view of the data:
```{r echo=TRUE}
str(data)
summary(data)
tail(data)
```
## What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
hist_data<-aggregate(data$steps, by=list(data$date), sum, na.rm=TRUE)
names(hist_data) = c("date", "steps")
hist(hist_data$steps,xlab="Total number of steps taken each day", breaks=20, col="grey", border="darkgrey", main="" )
```

2.Calculate and report the mean and median total number of steps taken per day
```{r echo=TRUE}
mean<-mean(hist_data$steps)
median<-median(hist_data$steps)
```
The mean of total number of steps taken per day is `r mean`. The median of total number of steps taken per day is `r median`

## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r echo=TRUE}
data$interval=as.factor(data$interval)
intvl_data=aggregate(data$steps, by=list(data$interval), mean, na.rm=TRUE)
names(intvl_data) = c("interval","avg.steps")
plot(1:288,intvl_data$avg.steps, xlab="Interval", ylab="Avg.steps/int", xlim=c(0,288), ylim=c(0,max(intvl_data$avg.steps)), xaxs="i", type="l", col="grey")
```


2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}
interval_max.steps = intvl_data[which.max(intvl_data$avg.steps),]
interval_max.steps
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

This was done by computing the number of rows from the original data frame that had NA values for the "steps" variable; in total, there are the following number of missing values:
```{r echo=TRUE}
nrow(data[is.na(data$steps)==TRUE,])
```

2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
&
3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

Using the "zoo" library and the "na.approx" function, we can fill in the missing values. The computation of the new dataset is presented below:

```{r echo=TRUE}
library(zoo)
new.data<-data

new.data$steps = na.approx(new.data$steps, na.rm=FALSE)
still_na= new.data[is.na(new.data$steps)==TRUE,]
still_na= match(still_na$interval, intvl_data$interval)
still_na = intvl_data$avg.steps[still_na]
new.data$steps[is.na(new.data$steps)==TRUE] = still_na

```
A brief overview of the newly created data:
```{r echo=TRUE}
str(new.data)
summary(new.data)
tail(new.data)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r echo=TRUE}
hist_data_new<-aggregate(new.data$steps, by=list(new.data$date), sum, na.rm=TRUE)
names(hist_data_new) = c("date", "steps")
hist(hist_data_new$steps,xlab="Total number of steps taken each day", breaks=20, col="grey", border="darkgrey", main="" )
```

```{r echo=TRUE}
mean_new<-mean(hist_data_new$steps)
median_new<-median(hist_data_new$steps)
```
The new values for mean and median are `r mean_new` and `r median_new`. The old values for mean and median (with NAs still existing in the data frame) were `r mean` and `r median`. Therefore, we can conlcude that the imputation of missing data has a substantial impact on mean and median values.

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r echo=TRUE}
new.data$type = weekdays(as.Date(new.data$date))
new.data$type = ifelse(new.data$type == "Saturday" | new.data$type == "Sunday", "weekend", "weekday")
new.data$type = as.factor(new.data$type)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r echo=TRUE}
new.data$interval= as.factor(new.data$interval)
new.data_avg= aggregate(new.data$steps, by=list(new.data$interval,new.data$type), mean, na.rm=TRUE)
names(new.data_avg) = c("interval", "type", "avg.steps")

library(lattice)
xyplot(avg.steps ~ interval | type, data=new.data_avg, layout=c(1,2), type="l", ylab="Average steps per time interval", col="grey")
```

Based on the graph above, it can be stated that weekends tend to have more evenly distributed number of steps/interval, but weekdays have higher average values of steps.
